Synthesizing Quotient Lenses

### 2018-03-15 papers/QuotientLenses/main.tex
sections 2 and 3 (and part of 4)

\begin{center}
  \begin{minipage}{2.2in}
    \centering
\begin{lstlisting}
@Book {Lovelace,
Author = "Ada Lovelace",
Title = {Notes},
}
\end{lstlisting}
  \end{minipage}
  \begin{minipage}{2in}
    \centering
\begin{lstlisting}
@Book{     Lovelace,
Title = {Notes},
Author = "Lovelace, Ada",     }
\end{lstlisting}
  \end{minipage}
\end{center}

When mapping these records into another format, such as EndNote, we
must decide what to do with the nonessential information.  A bijective
mapping must preserve all the information, including the extraneous
details, which leads to complex and brittle lenses complicated by the
need to preserve unimportant details.  
A better approach is to identify records that differ only in the
nonessential information, mapping them into a canonical representation.
This canonical representation is then mapped into the target format.
With this approach, both of the above \bibtex{} records would be
mapped to the same EndNote record.

@@

The permutation QRE combinator, \kw{perm}, allows data to be unordered. For example,
the following instance of \kw{perm} allows label, author and title fields
(which we assume have been defined earlier) to appear in any order.
%
\begin{lstlisting}
let bib_fields = perm (label, bib_author, bib_title)
\end{lstlisting}
%
To normalize the field separators, one can specify 
in an optional \kw{with} clause that the components of the
permutation are conjoined by another QRE. For instance, below, we normalize
whitespace between fields, leaving only a single newline.
%
\begin{lstlisting}
let bib_fields = perm (label, bib_author, bib_title) with (collapse ("," . wsp) $\to$ ",\n")
\end{lstlisting}

@@

Indeed this idea is what motivates our definition of {\em QRE lenses}.
Intuitively, our QRE lenses are bijective lenses with ``canonizers at the edges,''
just like Boomerang's quotient lenses. 
Figure~\ref{fig:attheedges} depicts the architecture of QRE lenses.
\begin{figure}[t]
\centering
\includegraphics[width=0.7\textwidth]{QREs}
\caption{QRE lens from source $P$ to target $Q$.  $P$ and $Q$ are QREs, each
  consisting of a ``whole'' set ($W(P)$ and $W(Q)$) and a ``kernel'' part
  ($K(P)$ and $K(Q)$) Between the two kernels is a plain bijective lens.
  The \kw{get} function of the whole lens takes an argument from $W(P)$,
  applies the canonizer for $P$ to obtain a canonical representative in
  $K(P)$, then applies the \kw{get} of the plain lens, yielding an element
  of $K(Q)$, which is a subset of $W(Q)$.  The \kw{put} function does the
  reverse, mapping from $W(Q)$ to $K(P)$ (hence $W(P)$.  This lens is in
  normal form, with canonizers (specified by QREs) at the outer edges and an
  ordinary bijective lens in the center. \BCP{Both the figure and the
    caption are new---what do you think??}}\saz{I think that this figure is a
  big improvement.  One tiny possible tweak: color the \kw{get} and \kw{put}
  keywords in the same color as in the text.}\bcp{Can do, if we keep the
  color in the text :-).  How are we going to decide about that?}\saz{Vote?}
\label{fig:attheedges}
\end{figure}
Every QRE lens $q_1$ has a type $q_1: Q_1 \Leftrightarrow Q_2$ where $Q_1$ and $Q_2$
are QREs. In the {\em get} direction, a QRE lens $q_1: Q_1 \Leftrightarrow Q_2$ uses the
source QRE $Q_1$ to compute a canonical representative for the data modulo the
equivalence relation defined by $Q_1$ and then applies the \kw{get} function of a
bijective string lens $\ell_1$ to this representative. In the {\em put} direction, $q_1$
operates similarly, but using the QRE $Q_2$ and the $\lput$ function of
$\ell_1$.

Because the QREs $Q_1$ and $Q_2$ determine the internal formats for data after
canonization, and because the algorithm for synthesizing bijective string lenses is
directed by these formats, $Q_1$ and $Q_2$ are all that is required to
synthesize QRE lenses end-to-end.

However, our restriction that canonizers appear only at the edges
raises a key technical question: Are we limiting the expressiveness of
our transformations by demanding all programs fit into this normal
form?  It turns out that we are not: any lens that uses canonizers
internally can be transformed into a lens that uses canonizers only at
the edges. The main technical contribution of this paper
(Theorem~\ref{normal form})
is a proof of this fact.
This technical result justifies 
using synthesis to produce QRE lenses instead of manually writing
them,
which can lead to substantial savings in program complexity.
For instance, after defining the \bibtex{} and EndNote QREs and
binding them to the variables \cd{bibtex} and \cd{endnote} respectively, then
all of the code in Figure~\ref{fig:example-lens} may be replaced by a single call
to the synthesis prodedure:
\begin{lstlisting}
let bib_to_end : bibtex $\Leftrightarrow$ endnote =
synth bibtex $\Leftrightarrow$ endnote using {(bib_example, end_example)}
\end{lstlisting}
\noindent 
Here, the generated quotient lens synchronizes \cd{bibtex} and
\cd{endnote} formats, using \cd{bib_example} and \cd{end_example} (the two
concrete example strings given at the beginning of this section) to
disambiguate. In addition, and as we saw earlier with the definition of
\cd{c_to_s}, the synthesis procedure itself can be used to create lenses that
are in turn used to define other QREs.  The ability to interleave QRE
specification with QRE lens synthesis yields a powerful and flexible way of
creating bidirectional transformations.

@@

The remaining QRE combinators, which we discuss next, provide simpler,
more compositional ways of building canonizers.  Nevertheless,
the \kw{normalize} combinator provides a useful guide in the design of
these combinators because it gives a sufficient condition for the
well-formedness of any potential QREs.

>>>>

\begin{center}
  \begin{minipage}{2.2in}
    \centering
\begin{verbatim}
@Book {Lovelace,
Author = "Ada Lovelace",
Title = {Notes},
}
\end{verbatim}
  \end{minipage}
  \begin{minipage}{2in}
    \centering
\begin{verbatim}
@Book{     Lovelace,
Title = {Notes},
Author = "Lovelace, Ada",     }
\end{verbatim}
  \end{minipage}
\end{center}

When mapping these records into another format, such as EndNote, we
must decide what to do with the nonessential information.  A bijective
mapping must preserve all the information, including the extraneous
details, which leads to complex and brittle lenses.
A better approach is to identify records that differ only in the
nonessential information, mapping them into a canonical representation.
This canonical representation is then mapped into the target format.
With this approach, both of the above \bibtex{} records would be
mapped to the same EndNote record.

@@

The permutation QRE combinator, \kw{perm}, allows data to be unordered. For example,
the following instance of \kw{perm} allows label, author, and title fields
(which we assume have been defined earlier) to appear in any order.
%
\begin{lstlisting}
let bib_fields = perm (label, bib_author, bib_title)
\end{lstlisting}
%
To normalize the field separators, one can specify 
in an optional \kw{with} clause that the components of the
permutation are conjoined by another QRE. For instance, below, we normalize
whitespace between fields, leaving only a single newline.
%
\begin{lstlisting}
let bib_fields = perm (label, bib_author, bib_title) with (collapse ("," . wsp) $\to$ ",\n")
\end{lstlisting}

@@

Indeed this idea is what motivates our definition of {\em QRE lenses}.
Intuitively, our QRE lenses are bijective lenses with ``canonizers at the edges,''
just like Boomerang's quotient lenses. 
Figure~\ref{fig:attheedges} depicts the architecture of QRE lenses.
\begin{figure}[t]
\centering
\includegraphics[width=0.7\textwidth]{QREs}
\caption{QRE lens from source $P$ to target $Q$.  $P$ and $Q$ are QREs, each
  consisting of a ``whole'' set ($W(P)$ and $W(Q)$) and a ``kernel'' part
  ($K(P)$ and $K(Q)$) Between the two kernels is a plain bijective lens.
  The \kw{get} function of the whole lens takes an argument from $W(P)$,
  applies the canonizer for $P$ to obtain a canonical representative in
  $K(P)$, then applies the \kw{get} of the plain lens, yielding an element
  of $K(Q)$, which is a subset of $W(Q)$.  The \kw{put} function does the
  reverse, mapping from $W(Q)$ to $K(P)$ (hence $W(P)$.  This lens is in
  normal form, with canonizers (specified by QREs) at the outer edges and an
  ordinary bijective lens in the center. \BCP{Both the figure and the
    caption are new---what do you think??}}\saz{I think that this figure is a
  big improvement.  One tiny possible tweak: color the \kw{get} and \kw{put}
  keywords in the same color as in the text.}\bcp{Can do, if we keep the
  color in the text :-).  How are we going to decide about that?}\saz{Vote?}
\label{fig:attheedges}
\end{figure}
Every QRE lens $q$ has a type $P \Leftrightarrow Q$ where $P$ and $Q$
are QREs. In the {\em get} direction, a QRE lens $q: P \Leftrightarrow Q$ uses the
source QRE $P$ to compute a canonical representative for the data modulo the
equivalence relation defined by $P$ and then applies the \kw{get} function of a
bijective string lens $\ell$ to this representative. In the {\em put} direction, $q$
operates similarly, but using the QRE $Q$ and the $\lput$ function of
$\ell$.

Because the QREs $P$ and $Q$ determine the internal formats for data after
canonization, and because the algorithm for synthesizing bijective string lenses is
directed by these formats, $P$ and $Q$ are all that is required to
synthesize QRE lenses end-to-end.

However, our requirement that canonizers appear only at the edges
raises a key technical question: Are we limiting the expressiveness of
our transformations by demanding all programs fit into this normal
form?  It turns out that we are not---any lens that uses canonizers
internally can be transformed into a lens that uses canonizers only at
the edges. The main technical contribution of this paper
(Theorem~\ref{normal form})
is a proof of this fact.
This technical result justifies 
using synthesis to produce QRE lenses instead of manually writing
them,
which can lead to substantial savings in program complexity.
For instance, after defining the \bibtex{} and EndNote QREs and
binding them to the variables \cd{bibtex} and \cd{endnote} respectively, then
all of the code in Figure~\ref{fig:example-lens} may be replaced by a single call
to the synthesis prodedure:
\begin{lstlisting}
let bib_to_end : bibtex $\Leftrightarrow$ endnote =
synth bibtex $\Leftrightarrow$ endnote using {(bib_example, end_example)}
\end{lstlisting}
\noindent 
Here, the generated quotient lens synchronizes \cd{bibtex} and
\cd{endnote} formats, using \cd{bib_example} and \cd{end_example} (the two
concrete example strings given at the beginning of this section) to
disambiguate. In addition, and as we saw earlier with the definition of
\cd{c_to_s}, the synthesis procedure itself can be used to create lenses that
are in turn used to define other QREs.  The ability to interleave QRE
specification with QRE lens synthesis yields a powerful and flexible way of
creating bidirectional transformations.

@@

The remaining QRE combinators, which we discuss next, provide simpler, more
compositional ways of building canonizers that meet these requirements by
construction.  Nevertheless, the \kw{normalize} combinator provides a useful
guide in the design of these combinators because it gives a sufficient condition
for the well-formedness of any potential QREs.
